Assessing landscape N removal in coastal New England catchments using the N-Sink

### 2023-05-12 manuscript/manuscript.Rmd
example ggplot for figure 4
Incorporate Ken's edits

---
title: Assessing landscape N removal in coastal New England catchments using the N-Sink
  approach with a new R package, nsink
author:
- Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\*^ ^*2*^, Arnold, C.^*3*^, Gold, A.^*1*^,
  Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
- ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*2*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
  CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
  Characterization and Remediation Division, Ada, OK 74820
- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water quality. Land management decisions made within estuarine watersheds have a direct impact on downstream N delivery. Natural features within watersheds can act as landscape sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous N, effectively removing it from the aquatic system. Identifying and evaluating these landscape sinks and their spatial relationship to N sources can assist managers to understand the effects of their decisions on downstream resources. N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 (or larger) catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment. A series of static maps are produced to visualize N removal efficiency, transport and delivery, the latter in the form of an index. To facilitate data acquisition, processing and visualization, an R package `nsink` was developed and has been used to produce static maps for all HUC-12 catchments in southern New England. The R package can be used to develop N delivery maps for a HUC-12 of interest, or users can visit the University of Connecticut website to explore the already mapped areas. Users can also investigate specific flowpaths interactively by clicking on any location within the catchment. A flowpath is generated, along with a table describing N removal along each segment.
---

@@

Excess nitrogen (N) delivery via surface water to coastal estuaries contributes to impaired water quality evidenced by excess algal blooms and hypoxia [@ryther1971nitrogen; @howarth2006nitrogen; @nixon1995coastal]. Hydrologic connections and flowpaths are influential in the delivery of nutrients to estuaries [@mengistu2020wetland]. Identifying landscape N sinks along hydrologic flowpaths -- areas where dissolved N is transformed to gaseous N and effectively removed from the aquatic system -- and their effect on N delivery at the watershed scale is helpful to watershed managers, land use planners and conservation organizations. The intent of N-Sink, and its associated R package `nsink`, is as a screening tool to identify source areas where downgradient N sinks are able to process and remove excess N, versus areas where downgradient N sinks are less prevalent or effective at removing N. The latter argues for more aggressive N management at the source, while the former argues for protecting and/or restoring downgradient N sinks that are helping to reduce N delivery to estuaries.

N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale (HUC = Hydrologic Unit Code; [@berelson2004mapping]), and makes extensive use of widely available GIS data. The theoretical underpinnings are outlined in Kellogg et al. [-@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [@mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds. Residence time in these landscape features is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].

The `nsink` R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps [@hollister2022nsink] . Datasets used by `nsink` are the National Hydrography Dataset Plus (NHDPlus), Soil Survey Geographic Database (SSURGO), the National Land Cover Dataset (NLCD) land cover and the National Land Cover Dataset (NLCD) impervious surface [@jin2019overall; @moore2019user; @soil2017web]. The static maps include 1) N Removal Efficiency, a color-coded map of the estimated N removal efficiency of the different landscape sinks; 2) N Transport Index, a heat map with the cumulative N removal along flowpaths originating from sources represented by a grid of points across a watershed. This heat map highlights “leaky” source areas with less downstream N retention versus those with higher downstream retention; and c) N Delivery Index, combining maps (1) and (2) along with source loading based on NLCD categories and expressed as an index ranging from 0 to 1, resulting in a map showing potential N delivery from different sources, taking into account the potential N removal as water moves downstream.

@@

## Development of the R package `nsink`
The package `nsink` implements the approach detailed in Kellogg et al. [-@kellogg2010geospatial] to estimate relative nitrogen (N) removal along a flowpath. The `nsink` package follows from an initial version of N-Sink written in ArcGIS using ModelBuilder. That version required time-consuming data manipulation by hand due to limitations of earlier NHDPlus datasets as well as the limitations of working in a GIS environment requiring a user license. With the increasingly more versatile GIS packages now available in R, the previously time-consuming spatial data acquisition and management of N-Sink was automated and applied to other HUC-12 catchments, leading to the development of the `nsink` package. Specifically, `nsink` relies on packages `sf` and `raster`. The `nsink` package is available from https://github.com/usepa/nsink and is fully described in [@hollister2022nsink].

@@

These static maps can be quickly plotted with `nsink_plot()` using the `map` argument to identify map type.

>>>>

---
title: Assessing landscape N removal in coastal New England catchments using the N-Sink
  approach with a new R package, nsink
author:
- Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\*^ ^*2*^, Arnold, C.^*3*^, Gold, A.^*1*^,
  Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
- ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*2*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
  CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
  Characterization and Remediation Division, Ada, OK 74820
- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: Excess nitrogen (N) loading to coastal ecosystems impairs estuarine water quality. Land management decisions made within estuarine watersheds have a direct impact on downstream N delivery. Natural features within watersheds can act as landscape sinks for N, such as wetlands, streams and ponds that transform dissolved N into gaseous N, effectively removing it from the aquatic system. Identifying and evaluating these landscape sinks and their spatial relationship to N sources can help managers understand the effects of alternative decisions on downstream resources. N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 (or larger) catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment. A series of static maps are produced to visualize N removal efficiency, transport and delivery, the latter in the form of an index. To facilitate data acquisition, processing and visualization, an R package `nsink` was developed and applied to produce static maps for all HUC-12 catchments in southern New England. The R package can be used to develop N delivery maps for a specific HUC-12 of interest, or users can visit the University of Connecticut website to explore the already mapped areas. Users can also investigate specific flowpaths interactively by clicking on any location within the catchment. A flowpath is generated, along with a table describing N removal along each segment.
---

@@

Excess nitrogen (N) delivery via surface water to coastal estuaries contributes to impaired water quality evidenced by excess algal blooms and hypoxia [@ryther1971nitrogen; @howarth2006nitrogen; @nixon1995coastal]. Hydrologic connections and flowpaths are influential in the delivery of nutrients to estuaries [@mengistu2020wetland]. Identifying landscape N sinks along hydrologic flowpaths -- areas where dissolved N is transformed to gaseous N and effectively removed from the aquatic system -- and their effect on N delivery at the watershed scale is helpful to watershed managers, land use planners and conservation organizations. N-Sink, and its associated R package `nsink`, were developed to be a screening tool that describes the areas or points within a watershed linked with flow paths where downgradient N sinks process and remove excess N, versus areas where downgradient N sinks are less prevalent or less effective at removing N. This watershed flow path approach provides a means to easily visualize the relative sensitivity of an area within a watershed that may require more aggressive N management at the source, protection of existing N retentive areas, or restoration and construction of downgradient N retentive areas that are needed to reduce N delivery to estuaries.

N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale (HUC = Hydrologic Unit Code; [@berelson2004mapping]), and makes extensive use of widely available GIS data. The theoretical underpinnings are outlined in Kellogg et al. [-@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [@mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds. Residence time within landscape features that support N removal is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].

The `nsink` R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps [@hollister2022nsink] . Datasets used by `nsink` include the National Hydrography Dataset Plus (NHDPlus), Soil Survey Geographic Database (SSURGO), the National Land Cover Dataset (NLCD) land cover and the National Land Cover Dataset (NLCD) impervious surface [@jin2019overall; @moore2019user; @soil2017web]. The static maps illustrate 1) N Removal Efficiency, a color-coded map of the estimated N removal efficiency of the different landscape sinks; 2) N Transport Index, a heat map with the cumulative N removal along flowpaths originating from sources represented by a grid of points across a watershed. This heat map highlights “leaky” source areas with less downstream N retention versus those with higher downstream retention; and c) N Delivery Index, combining maps (1) and (2) along with source loading based on NLCD categories and expressed as an index ranging from 0 to 1, resulting in a map showing potential N delivery from different sources, taking into account the potential N removal as water moves downstream. N-Sink and `nsink` integrate these data to make the application of complex hydrologic and biogeochemical processes accessible to end users and stakeholders in a visually clear and useful format. Here we describe the development of the package, and example workflow, and simple use applications to demonstrate the capabilities of the tools.

@@

## Development of the R package `nsink`
The package `nsink` implements the approach detailed in Kellogg et al. [-@kellogg2010geospatial] to estimate relative nitrogen (N) removal along a flowpath. The `nsink` package follows from an initial version of N-Sink written in ArcGIS using ModelBuilder. The initial version required time-consuming data manipulation by hand due to limitations of earlier NHDPlus datasets as well as the limitations of working in a GIS environment requiring a user license. With the increasingly more versatile GIS packages now available in R, the previously time-consuming spatial data acquisition and management of N-Sink was automated and applied to other HUC-12 catchments, leading to the development of the `nsink` package. Specifically, `nsink` relies on packages `sf` and `raster`. The `nsink` package is available from https://github.com/usepa/nsink and is fully described in [@hollister2022nsink].

@@

These static maps can be quickly plotted with `nsink_plot()` using the `map` argument to identify map type.  These plots are designed for exploration and not necessarily for high quality maps.  For higher quality maps, the source data can be plotted in software of a users choice (e.g. QGIS, `ggplot2`).
